---
title: "Reporting"
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

The below reports data quality and outcomes.
See [manuscript.qmd](/report/manuscript.qmd) for draft text (or render to markdown: `quarto::quarto_render(here::here("report", "manuscript.qmd"))`).

```{r}
#| label: setup

library(here)
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(scales)
library(forcats)
library(lubridate)
library(viridis)
library(plotly)

data_processed <- read_csv(here("data", "processed", "participants.csv"))
participants <- data_processed |>
  filter(record_remove == "include", !is.na(weight))
exclusions <- read_csv(here("data", "processed", "data-quality.csv"))
raw_total   <- sum(exclusions$n_records)
final_inc   <- exclusions |> filter(record_remove == "include")
n_included_records      <- final_inc$n_records
n_included_participants <- final_inc$n_participants
```

### Data processing

- Weight data come from `r raw_total` raw records across two ODK survey forms
  - `form1`: baseline enrolment record with all characteristics
  - `form2`: follow-up (date, weight measurement only)
- `R/clean.R` combines both forms into a single long dataset, `data/processed/participants.csv`
- All collected records are included in the clean csv, including invalid records or implausible values
  - Records are assessed against each exclusion criterion in turn, as below
  - Reasons for exclusion are flagged in the field `record_remove`
  - Records passing all criteria are labelled `include`

#### Exclusions

```{r}
#| label: exclusion-table

exclusions |>
  transmute(
    `Record processing`   = reason,
    `Label`            = record_remove,
    `Records`          = n_records,
    `Participants`     = n_participants,
    `Records remaining` = n_remaining,
    `Participants remaining` = n_participants_remaining
  ) |>
  kable()
```

Note around 200 records in `form2` couldn't be matched to any participant ID in `form1` (flagged as `invalid_no_baseline`).
As a result these records have no pre-war weight, height, or demographics.

- These might be participants who only ever entered data to the follow-up form
- Or it might be participants who enrolled at baseline but used a different ID combination to enter the follow up form
- The follow-up form has no other identifiers, so no obvious way to identify this

### Data description

```{r}
#| label: plot-setup

theme_set(theme_bw() + theme(legend.position = "bottom"))

bmi_levels <- c("Underweight", "Normal", "Overweight", "Obese")
bmi_palette <- setNames(viridis(length(bmi_levels), direction = -1), bmi_levels)
age_levels <- c("Age under 30", "Age 30-45", "Age over 45")

# derive follow-up flag, ordered age
included <- participants |>
  add_count(id, name = "n_records") |>
  mutate(
    age = factor(age, levels = age_levels),
    has_followup = n_records >= 2
  )

# baseline characteristics
baseline <- included |> filter(record_number == 1)
```

#### Participant characteristics

```{r}
#| label: tbl-table1
#| tbl-cap: "Characteristics of participants at study entry, overall and by whether they later contributed follow-up measurements"

library(gtsummary)

baseline |>
  mutate(group = if_else(has_followup, "Contributed follow-up", "Baseline only")) |>
  select(group,
         `BMI at entry (kg/m2), median [IQR]` = bmi_entry,
         Sex                    = sex,
         `Age group`            = age,
         Governorate            = governorate,
         Role                   = role,
         Organisation           = organisation) |>
  tbl_summary(
    by = group,
    type = all_continuous() ~ "continuous",
    statistic = list(all_continuous()  ~ "{median} [{p25}-{p75}]",
                     all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    missing_text = "(Missing)"
  ) |>
  add_overall() |>
  as_kable()
```

```{r}
#| label: fig-participation-over-time
#| fig-cap: "Weekly reported weight measurements, by new participants (baseline records) and the returning cohort (follow-up records)."

participation_weekly <- included |>
  mutate(
    week = floor_date(as.Date(record_date), "week"),
    source = if_else(record_number == 1, "New participant", "Follow-up") |>
      factor(levels = c("Follow-up", "New participant"))
  ) |>
  count(week, source, name = "n_measurements")

plot <- ggplot(participation_weekly, aes(x = week, y = n_measurements, fill = source)) +
  geom_col(width = 6) +
  scale_fill_viridis_d(direction = -1, name = NULL) +
  labs(x = NULL, y = "Reported measurements per week")
plot
```

#### Participant outcomes

##### Weight pre-war to study entry

```{r}
#| label: fig-prewar-pct-change
#| fig-cap: "Participant weight change from pre-war to study entry."

plot <- baseline |>
  filter(!is.na(weight_prewar), !is.na(weight_entry)) |>
  mutate(pct_change = (weight_entry - weight_prewar) / weight_prewar * 100) |>
  ggplot(aes(x = pct_change)) +
  geom_histogram(binwidth = 1) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_fill_viridis() +
  labs(x = "Weight change pre-war to study entry (% kg)", y = "Participants")
plot
```

```{r}
#| label: fig-age-sex-pyramid
#| fig-cap: "Age-sex distribution by BMI category, pre-war and at study entry."

plot <- baseline |>
  filter(sex %in% c("Male", "Female"), !is.na(age)) |>
  select(id, sex, age, `Pre-war` = bmi_category_prewar, `Study entry` = bmi_category) |>
  pivot_longer(c(`Pre-war`, `Study entry`),
               names_to = "period", values_to = "bmi_cat") |>
  filter(!is.na(bmi_cat)) |>
  mutate(period  = factor(period, levels = c("Pre-war", "Study entry")),
         bmi_cat = factor(bmi_cat, levels = bmi_levels)) |>
  count(period, sex, age, bmi_cat) |>
  mutate(n_signed = if_else(sex == "Male", -n, n)) |>
  ggplot(aes(x = n_signed, y = age, fill = bmi_cat)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "#01454f", linewidth = 0.3) +
  scale_x_continuous(labels = abs) +
  scale_fill_manual(values = bmi_palette, name = "BMI category") +
  facet_wrap(~period) +
  labs(x = "Participants  (Male ← | → Female)", y = NULL)
plot
```

##### Study entry through follow-up

```{r}
#| label: fig-weight-trajectories-calendar
#| fig-cap: "Individual trajectories by calendar date among those with any recorded follow-up"

plot <- included |>
  filter(has_followup) |>
  ggplot(aes(x = record_date, y = bmi)) +
  geom_line(aes(group = id), alpha = 0.3, colour = "#01454f") +
  geom_point(alpha = 0.3, size = 1, colour = "#01454f") +
  geom_smooth(method = "loess", se = TRUE,
              colour = "#ffa600", fill = "#ffa600", alpha = 0.15) +
  labs(x = "Date of measurement", y = "BMI")
plot
```
